\chapter{Principal Component Analysis }
\label{ch:pca}


Machine learning is the field of study that gives computers the ability to learn from data without being explicitly programmed. In this context, a model is a mathematical function or computational structure that transforms input data into an output, following rules that are not explicitly defined but inferred from examples. The data are represented through numerical features, and the model uses them to learn relationships, trends or structures that are relevant for a given task, such as classification or dimensionality reduction. During training, the model adjusts its internal parameters according to the examples it receives, in order to capture patterns that remain informative when new observations are presented.

This approach is well suited to Raman spectroscopy of biological tissues. As discussed in \Cref{ch:introduction}, the complexity of biological Raman spectra makes it difficult to reduce the classification problem to a few manually selected bands. For this reason, the entire spectrum is treated as a numerical object whose overall profile may contain information relevant to the classification task. In the dataset used in this work, the Raman measurements are already organized in a computationally suitable tabular form. Starting from the structure described in \Cref{ch:Dataset}, the input data for the machine learning analysis are obtained from the spectral columns \texttt{band\_}n. These columns contain the Raman intensity values at the 483 selected shifts; from a machine learning perspective, they form the feature vector associated with each pixel-level spectrum acquired from a Raman map. When all spectra are considered together, the corresponding feature vectors are arranged into a data matrix, whose rows represent spectra and whose columns represent Raman-shift variables. 

Machine learning methods are commonly distinguished according to the information available during training. When the aim is to train a classifier, the learning process is supervised: the model is shown both the input data and the correct output associated with each example. In this way, it can learn how spectral information is related to tissue classes and use this relation to assign a class to new spectra. In the present work, this corresponds to using each Raman spectrum as input and its associated tissue label as the target output for binary classification.

When labels are not part of the learning process, the analysis becomes unsupervised: the algorithm only receives the input data and searches for structure within them. This approach is useful when the goal is not to assign a class directly, but to explore the organization of the data, visualize similarities between spectra or reduce the number of variables. PCA belongs to this second case. 

This chapter focuses on Principal Component Analysis (PCA), an unsupervised 
dimensionality-reduction method that identifies the directions along which the spectra show the largest variance. PCA projects the original spectral variables onto a new set of components, providing a more compact representation of the data. This transformation is useful because dimensionality reduction can be performed by retaining only the first principal components, which account for most of the spectral variance, and discarding later components associated with smaller contributions. In this way, each spectrum is represented by a reduced set of PCA scores instead of the full set of original Raman-shift variables. This compact representation can reduce redundancy and may limit the influence of weak or noisy variations before classification.

In this work, PCA is therefore first used as an
exploratory analysis of the dataset. Its theoretical formulation is presented with particular attention to the identification of the principal components, the amount of variance retained after projection, and the interpretation of the spectra in the transformed PCA space. When PCA is later used inside the classification pipeline, however, it must be fitted only on the training data of each split or fold and then applied to the corresponding validation or test data. This preserves the separation between training and evaluation data, preventing information from the validation or test set from influencing the learned representation.

\section{Dimensionality reduction}
\label{sec:dimensionality_reduction}

The spectral dataset can be represented as a numerical matrix
\begin{equation}
    \mathbf{X} \in \mathbb{R}^{n \times p},
    \label{eq:pca_data_matrix}
\end{equation}
where \(n\) is the number of spectra and \(p\) is the number of spectral variables. In the
dataset described in \Cref{ch:Dataset}, each spectrum is represented by \(p = 483\)
Raman-shift intensity values after the removal of the silent region. Thus, each spectrum
can be interpreted not only as a curve, but also as a point in a \(483\)-dimensional
feature space, where each coordinate corresponds to the intensity measured at one Raman
shift.

Working directly with all the original variables is possible, but it is not always the
most convenient representation. In Raman spectra, nearby variables are often correlated because spectral bands are broad and can overlap. Therefore, the \(483\) spectral variables do not necessarily correspond
to \(483\) independent sources of information. Dimensionality reduction addresses this
situation by replacing the original variables with a smaller number of new variables
designed to retain the main structure of the dataset.

The motivation for this step is also connected to the so-called curse of dimensionality.
In low-dimensional spaces, data points can be relatively close to one another, so local
neighbourhoods and similarities are easier to identify. When the number of dimensions
increases, the volume of the feature space grows exponentially with the number of
variables. Therefore, keeping the same density of samples would require an exponentially
larger number of training observations. If the number of available spectra remains fixed,
the dataset becomes sparse: the points are spread across a much larger space and tend to
be far from one another. This has an important consequence for learning algorithms: if a
new observation is far from the training examples, the model has less direct evidence on
which to base its prediction, and the risk of learning patterns that are specific to the
training set increases. 

This does not mean that dimensionality reduction is automatically beneficial. Removing
dimensions inevitably removes part of the original information, and the discarded
variation may include subtle class-related spectral differences. For this reason, dimensionality reduction does not necessarily lead to an improvement. Its effect must be interpreted together with the downstream
classification performance and with the amount of variance retained in the reduced
representation.

One common way to reduce dimensionality is projection. In projection-based methods, the
original data are represented in a lower-dimensional subspace by projecting each sample
onto a new set of axes. These axes define new variables, which are combinations of the
original features. If the data lie close to a lower-dimensional linear subspace, this
projection can preserve much of the relevant structure while using fewer coordinates. PCA
belongs to this projection-based family of methods.

\section{Principal Component Analysis}
\label{sec:pca_theory}

Principal Component Analysis (PCA) searches for a new orthogonal coordinate system in
which the variability of the data is represented as efficiently as possible. Starting from
centered observations, the first principal component is chosen as the direction along which
the projected data have maximum variance. The second component describes the largest
remaining variance under the constraint of being orthogonal to the first, and the same
criterion is applied to the following components. In this way, PCA orders the new axes from
the most to the least informative in terms of variance.

Since PCA is unsupervised, class labels are not involved in the computation of these axes.
The method therefore identifies directions that summarize the dominant variability of the
spectra, not necessarily the directions that best separate tissue classes. For this reason,
when PCA is used before a classifier, the usefulness of the reduced representation must be
assessed through the performance of the complete classification pipeline.

\subsection{Data centering}
\label{subsec:pca_centering_covariance}

Let \(\mathbf{X} \in \mathbb{R}^{n \times p}\) be the data matrix, where each row
\(\mathbf{x}_i^{T}\) is one spectrum and each column is one spectral variable. PCA is
applied to centered data. The mean vector is
\begin{equation}
    \bar{\mathbf{x}} =
    \frac{1}{n}\sum_{i=1}^{n}\mathbf{x}_i,
    \label{eq:pca_mean_spectrum}
\end{equation}
the centered data matrix is
\begin{equation}
    \mathbf{X}_c =
    \mathbf{X} - \mathbf{1}\bar{\mathbf{x}}^{T},
    \label{eq:pca_centered_matrix}
\end{equation}
where \(\mathbf{1}\) is a column vector of ones. Centering subtracts the average spectrum
from each observation, so PCA describes variations around the mean profile rather than
absolute intensity levels.

\subsection{Singular Value Decomposition}
\label{subsec:pca_svd}

PCA can be obtained through
the Singular Value Decomposition (SVD) of the centered data matrix. The SVD is a stable
matrix factorization that represents the data in terms of directions adapted to the dataset
itself. Applied to \(\mathbf{X}_c\), it writes
\begin{equation}
    \mathbf{X}_c =
    \mathbf{U}\boldsymbol{\Sigma}\mathbf{V}^{T},
    \label{eq:pca_svd}
\end{equation}
where \(\mathbf{U}\) and \(\mathbf{V}\) have orthonormal columns and
\(\boldsymbol{\Sigma}\) is diagonal.  
The matrix \(\mathbf{V}\) contains the principal components \(\mathrm{PC}_j\):
\begin{equation}
    \mathbf{V} =
    \begin{bmatrix}
    \mathrm{PC}_1 & \mathrm{PC}_2 & \cdots & \mathrm{PC}_p
    \end{bmatrix}
    \in \mathbb{R}^{p \times p}.
    \label{eq:pca_v_matrix}
\end{equation}
 The columns of
\(\mathbf{V}\) are the right singular vectors; each of them specifies
how the original Raman-shift variables are weighted to form one principal component.
In the PCA terminology, these are the loading vectors.

The diagonal entries of \(\boldsymbol{\Sigma}\) are the singular values:
\begin{equation}
    \boldsymbol{\Sigma} =
    \begin{bmatrix}
    \sigma_1 & 0 & \cdots & 0 \\
    0 & \sigma_2 & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \cdots & \sigma_p
    \end{bmatrix}
    \in \mathbb{R}^{p \times p}.
    \label{eq:pca_sigma_matrix}
\end{equation}
These values indicate the relative importance of the corresponding components in
describing the variability of the dataset. Since they are ordered from largest to smallest,
the SVD naturally provides a hierarchy of components, from the dominant patterns to
progressively weaker contributions.

The columns of \(\mathbf{U}\) are the corresponding left singular vectors, which provide
normalized coordinates of the spectra along these directions:
\begin{equation}
    \mathbf{U} =
    \begin{bmatrix}
    \mathbf{u}_1 & \mathbf{u}_2 & \cdots & \mathbf{u}_p
    \end{bmatrix}
    \in \mathbb{R}^{n \times p}.
    \label{eq:pca_u_matrix}
\end{equation}
When these coordinates are scaled by the singular values in \(\boldsymbol{\Sigma}\), they
give the projection of the original spectra onto the principal components. These projected
coordinates are the PCA scores.

This computational approach has a statistical interpretation in terms of variance and
covariance. Variance measures the spread of the data along a direction, while covariance
describes how spectral variables vary together. PCA is linked to these quantities because
it searches for directions obtained as combinations of the original variables, choosing
directions along which the projected data show large variability.

Let \(\mathbf{W}_d\) be the matrix containing the first \(d\) columns of
\(\mathbf{V}\). Following the notation used for projection, \(\mathbf{W}_d\) contains the
first \(d\) principal directions. For \(d\) retained components, the projected data matrix
is
\begin{equation}
    \mathbf{X}_{d\text{-proj}} =
    \mathbf{X}_c \mathbf{W}_d =
    \mathbf{U}_d\boldsymbol{\Sigma}_d,
    \label{eq:pca_scores_svd}
\end{equation}
where \(\mathbf{X}_{d\text{-proj}} \in \mathbb{R}^{n \times d}\). Each row corresponds
to one spectrum expressed in the PCA space.

Keeping only the first \(d\) directions gives a rank-\(d\) approximation of the centered
data:
\begin{equation}
    \mathbf{X}_c \approx
    \mathbf{U}_d\boldsymbol{\Sigma}_d\mathbf{W}_d^{T}.
    \label{eq:pca_low_rank}
\end{equation}
This truncated representation is the basis of dimensionality reduction in PCA: each
spectrum is described by \(d\) scores instead of all \(p\) original variables. The choice
of \(d\) is guided by the explained variance. Each principal component accounts for a
fraction of the total variability in the data; the cumulative explained variance is the
progressive sum of these fractions up to the \(d\)-th component and indicates
how much of the original variability is retained after projection.

 

\subsection{Computational implementation}
\label{subsec:pca_computational_implementation}

In Python, PCA can be implemented using the \texttt{PCA} class provided by
scikit-learn. The input is a numerical data matrix \(\mathbf{X}\), whose rows are the
observations and whose columns are the original variables. Once the PCA object is fitted,
the method \texttt{transform} projects the data onto the principal directions. The
combined method \texttt{fit\_transform} performs both operations in a single call:

\begin{verbatim}
from sklearn.decomposition import PCA

pca = PCA(n_components=d)
X_d_proj = pca.fit_transform(X)
\end{verbatim}

In this code, \texttt{n\_components} is the number \(d\) of retained principal
components, and \texttt{X\_d\_proj} is the projected data matrix. Scikit-learn centers
the data internally before computing PCA, so the user provides the original data matrix
and the algorithm estimates the principal axes from it.

When the number of components is not known in advance, PCA can first be fitted without
specifying \texttt{n\_components}. The cumulative explained variance can then be used as
a diagnostic to choose a possible value of \(d\):


\begin{verbatim}
import numpy as np

pca = PCA()
pca.fit(X)

cumsum = np.cumsum(pca.explained_variance_ratio_)
d = np.argmax(cumsum >= 0.95) + 1
\end{verbatim}

The attribute \texttt{explained\_variance\_ratio\_} contains the fraction of variance
associated with each principal component. Its cumulative sum indicates how much of the
total variance is retained when the first components are kept. In the example above,
\(d\) is the smallest number of components required to preserve at least \(95\%\) of the
variance. This threshold is only a practical criterion: the final number of components
should still be evaluated with respect to the goal of the analysis or of the predictive
model.




\section{PCA analysis of the VIBRA dataset}