
\chapter{Aim of the thesis}
\label{sec:aim_thesis}

The present work is based on spontaneous Raman hyperspectral maps acquired from
patient-derived tumour slice cultures of head and neck squamous cell carcinoma
(HNSCC). These samples are ex vivo tissue sections obtained from surgical tumour
specimens and preserve, at least in part, the morphological and biochemical heterogeneity
of the original tissue. 

The acquisition of the Raman maps and the preprocessing of the raw
spectra were not performed as part of this thesis; the analysis starts from spectra already
calibrated, preprocessed and organized in a form suitable for computational processing.

***(la parte sopra potrebbe sostituire completamente il paragrafo su data provenence?)***


Each measured point of a Raman map is associated with a complete Raman spectrum and
with a spatial position within the tissue section. The available data therefore combine
spectral information, related to the biochemical composition of the measured region, and
spatial information, related to the arrangement of the sampled points within the map.

The aim of this thesis is to evaluate a machine learning approach for the binary
classification of Raman spectra, with particular attention to the recognition of tumour
tissue. The analysis combines an exploratory study based on Principal Component
Analysis (PCA) with the implementation of an artificial neural network (ANN) classifier.
Two binary classification settings are considered: the discrimination between tumour and
stroma, and the discrimination between tumour and all non-tumour tissue classes included
in the available annotations.

The work will be developed through three main computational steps:
\begin{itemize}
    \item an exploratory analysis of the available dataset;
    \item the application of Principal Component Analysis (PCA) to inspect the spectral
    variability and obtain a reduced representation of the spectra;
    \item the implementation of an artificial neural network (ANN) for binary
    classification.
\end{itemize}

necessario dire questa cosa?--->
***The results should be interpreted as an assessment of this computational pipeline on
the available Raman dataset. They do not constitute a clinical validation of Raman
spectroscopy as a diagnostic tool. Such a validation would require a larger and
independent cohort of samples, a predefined clinical protocol and systematic comparison
with established diagnostic procedures.***



