\chapter{Materials and methods}
\label{ch:materials_methods}

\section{Data provenance}
\label{sec:data_provenance}

The Raman data used in this project were not acquired as part of the present thesis.
The analyses start from spectra that had already been collected, calibrated and
preprocessed before being provided in dataset form. 
\subsection{Experimental setup}
\label{sec:experimental_setup}

The measurements were performed with a home-built Raman microspectroscopy system
developed for biological samples. The excitation source is a continuous-wave
diode-pumped laser operating in the red spectral region, around \(660~\mathrm{nm}\),
providing the monochromatic radiation required for Raman measurements.
The Raman scattering efficiency increases strongly at shorter wavelengths, approximately as
\begin{equation}
    I_R \propto \frac{1}{\lambda_0^4},
    \label{eq:lambda_dependence}
\end{equation}
where \(\lambda_0\) is the excitation wavelength.
Shorter wavelengths can therefore increase the Raman signal and can also improve spatial resolution. However, they may also increase the risk of photodamage and can enhance unwanted fluorescence background. A red excitation wavelength provides a practical balance between signal intensity, spatial resolution and sample preservation.

Along the excitation path, the laser beam is first filtered to remove unwanted spectral
components and then attenuated by neutral-density filters. The beam is then expanded by
a \(5\times\) beam expander, from an initial diameter of about \(0.7~\mathrm{mm}\) to
about \(3.5~\mathrm{mm}\), in order to better match the back aperture of the microscope
objective and ensure proper focusing on the sample plane.

Inside the inverted microscope, the excitation beam is focused onto the tissue section
using a \(50\times\) dry objective with numerical aperture \(\mathrm{NA}=0.8\) and
working distance of \(1~\mathrm{mm}\). The same objective collects the backscattered
light from each illuminated region, while the motorized stage scans the sample point by
point to associate each spatial position with a Raman spectrum.

The collected light contains both the weak Raman signal and a much stronger Rayleigh
component with the same wavelength as the excitation laser. Since Rayleigh scattering is
several orders of magnitude more intense than Raman scattering, it must be rejected
before detection, because it could hide the Raman signal and saturate the detector. In
the detection path, the dichroic beamsplitter separates these contributions, while a
\(664~\mathrm{nm}\) long-pass edge filter suppresses residual Rayleigh components
before detection.

After filtering, the Raman signal is focused by a lens into the spectrometer, where it is
dispersed by a diffraction grating. A \(600~\mathrm{grooves/mm}\) grating is typically
used to cover both the fingerprint and the CH regions. The dispersed light
is recorded by a CCD detector with a \(1024\times256\) pixel sensor. Vertical binning
along the \(256\)-pixel axis is used, so that the raw spectral information is recorded
along the \(1024\)-pixel horizontal axis. Through calibration, this pixel axis is
converted into Raman shift, producing the final spectrum expressed as intensity versus
Raman shift.

For Raman hyperspectral maps, spectra were acquired point by point with a spatial step
of \(5~\mu\mathrm{m}\), using an exposure time of \(500~\mathrm{ms}\) and a laser power
of \(120~\mathrm{mW}\) at the source. These acquisition parameters reflect a compromise
between signal quality, acquisition time and preservation of the biological specimen.

\subsection{Preprocessing}
\label{sec:preprocessing_before_analysis}

The raw spectral
data underwent a preprocessing workflow aimed at making the spectra comparable and
suitable for computational analysis. 

The detector pixel axis was calibrated and converted
into Raman shift, and the spectra were interpolated onto a common wavenumber axis.

Unwanted experimental contributions were then corrected or removed, including
spike-like artifacts due to cosmic rays, the quartz substrate/background contribution and
baseline distortions. Normalization steps were applied where required to reduce intensity
variations not directly related to tissue composition.

The silent region \(1800\)--\(2800~\mathrm{cm}^{-1}\), where biological Raman spectra usually contain few relevant bands, was excluded. After these
operations, each spectrum is represented by \(483\)
Raman-shift variables spanning the fingerprint and CH regions.

dirlo qui o nella sezione sul dataset?
***Non-informative
or unreliable pixels, such as transparent regions and acquisition artifacts, were removed
from the computational dataset. ***

This preparation is important for machine learning, because the spectra used as input to
the classification model are shaped not only by tissue biochemistry, but also by
acquisition conditions and experimental noise. If these effects are not properly
controlled or corrected, the model may learn patterns related to measurement variability
rather than tissue class.

\section{Dataset}
